Evaluation of LOOCV predictive abilities

### 2016-09-21 reports/agronomic_data.Rmd
ss-pred report, single doc with toc

## Skewness
A general rule of thumb to consider is that skewed data whose ratio of the 
highest value to the lowest value is greater than 20 have significant skewness.
The formula for the skewness statistic is
$$
\text{skewness}  = \frac{\sum (x_i - \bar{x})^{3}}{(n - 1)v^{3/2}},
$$
where $v = \frac{\sum (x_i - \bar{x})^2}{(n - 1)}$ with $x$ is the predictor
variable, $n$ is the number of values and $\bar{x}$ is the sample mean of the 
predictor.
```{r Agro-Skewness, echo = FALSE}
skew_tab_cap <- paste(
  "Skewness of seven agronomic traits for the full set of hybrids and the",
  "reduced set of hybrids, respectively."
)
all_skewness <- all_pheno %>%
  group_by(Trait) %>%
  summarize(Skewness = e1071::skewness(Observed),
            Skewness = round(Skewness, digits = 2))
reduced_skewness <- reduced_pheno %>%
  group_by(Trait) %>%
  summarize(Skewness = e1071::skewness(Observed),
            Skewness = round(Skewness, digits = 2))
full_join(x = all_skewness, y = reduced_skewness,
          by = "Trait",
          suffix = c(" (Full)", " (Reduced)")) %>%
  kable(caption = skew_tab_cap, booktable = TRUE)
```

## Distribution
```{r Agro-Distribution, message = FALSE, echo = FALSE, fig.cap = "Histogram of the distribution of seven agronomic traits. Black bins show the distribution of the reduced set of hybrids whereas light grey bins show the distribtution of the full set of hybrids."}
ggplot(all_pheno, aes(x = Observed)) +
  geom_histogram(fill = "grey", alpha = 0.5) +
  geom_histogram(data = reduced_pheno, color = "black") +
  facet_wrap(~ Trait, scales = "free") +
  theme_bw()
```

>>>>

## Skewness
A general rule of thumb to consider is that skewed data whose ratio of the 
highest value to the lowest value is greater than 20 have significant skewness.
The formula for the skewness statistic is
$$
\text{skewness}  = \frac{\sum (x_i - \bar{x})^{3}}{(n - 1)v^{3/2}},
$$
where $v = \frac{\sum (x_i - \bar{x})^2}{(n - 1)}$ with $x$ denoting the 
predictor variable, $n$ denoting the number of values and $\bar{x}$ denoting the 
sample mean of the predictor.
```{r Agro-Skewness, echo = FALSE}
skew_tab_cap <- paste(
  "Skewness of seven agronomic traits for the full set of hybrids and the",
  "reduced set of hybrids, respectively."
)
all_skewness <- all_pheno %>%
  group_by(Trait) %>%
  summarize(Skewness = e1071::skewness(Observed),
            Skewness = round(Skewness, digits = 2))
reduced_skewness <- reduced_pheno %>%
  group_by(Trait) %>%
  summarize(Skewness = e1071::skewness(Observed),
            Skewness = round(Skewness, digits = 2))
full_join(x = all_skewness, y = reduced_skewness,
          by = "Trait",
          suffix = c(" (Full)", " (Reduced)")) %>%
  kable(caption = skew_tab_cap, booktabs = TRUE)
```

## Distribution
```{r Agro-Distribution, message = FALSE, echo = FALSE, fig.cap = "Histogram of the distribution of seven agronomic traits. Black bins show the distribution of the reduced set of hybrids whereas light grey bins show the distribution of the full set of hybrids."}
ggplot(all_pheno, aes(x = Observed)) +
  geom_histogram(fill = "grey", alpha = 0.5) +
  geom_histogram(data = reduced_pheno, color = "black") +
  facet_wrap(~ Trait, scales = "free") +
  theme_bw()
```

### 2016-09-21 reports/index.Rmd
ss-pred report, single doc with toc

---
title: Single-step Hybrid Predition
author: Matthias Westhues, Claas Heuer
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
output:
  bookdown::gitbook:
documentclass: book
bibliography: ["cheuer.bib"]
biblio-style: apalike
link-citations: yes
---

>>>>

---
title: Single-step Hybrid Predition
author: Matthias Westhues, Claas Heuer
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
documentclass: book
bibliography: ["library.bib", "cheuer.bib"]
biblio-style: apalike
link-citations: yes
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

### 2016-09-21 reports/intro.Rmd
ss-pred report, single doc with toc

# Introduction
The integration of different sources of information for the prediction of 
breeding or production values is one of the major challenges in current genomic 
prediction research.
In the early years of genomic prediction, a 2-step procedure was the most 
intuitive choice for establishing a new selection scheme.
The first consists of conventionally estimating breeding values based on 
pedigree relationships and therefore the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (i.e. high 
number of offspring), enter the second step in which SNP genotypes are regressed 
against the conventional breeding values.
Those markers can then be used to predict genomic breeding values for solely 
genotyped and potentially very young individuals.
Considering the two sources of information &mdash; pedigree and high density SNP 
genotypes &mdash; the process can be described as follows:

@@

where $\mathbf{X}$ is a design matrix for fixed effects, $\mathbf{b}$ is a 
vector of fixed effects, $\mathbf{Z}$ is the design matrix for the random animal 
effects, which associate observations in $\mathbf{y}$ with elements in 
$\mathbf{A}$, $\mathbf{a}$ is the vector of random animal effects (=breeding 
values) and $\mathbf{e}$ is a vector of iid residual errors with expectation 
$0$ and variance $\sigma^2_e$.
The distribution of the random vector $\mathbf{u}$ is 
$MVN(0, \mathbf{A}\sigma^2_a)$.

### Second Step
In the second step, a subset of the individuals for which conventional breeding 
values have been estimated, is selected and enters the reference population for 
training a genomic prediction model.
That selection requires that the individual is genotyped and has 
$\text{PEV}(u_{\text{candidate}}) \leq \text{threshold}$.
As the breeding values from step one enter the genomic prediction model as 
phenotypes, one has to be careful about the residual error distribution of
such a model.
For instance, if the range of the prediction error variances is very broad in 
the reference population, the residual errors will not be identically 
distributed anymore, e.g. they should be weighted differently or estimated 
individually.
One popular approach to account for that problem is *Deregressing* the breeding 
values and weighting the residuals according to the prediction error variances 
[@Garrick2009].
Genomic breeding values for any animal in the population can then be obtained 
by:

$$
\begin{equation}
	\label{eq:gebv}
	\mathbf{\hat{g}} =  (\mathbf{M} - 2\mathbf{p})\mathbf{\hat{u}},
\end{equation}
$$

where $\mathbf{p}$ is a row-vector of allele frequencies for all the markers in 
$\mathbf{M}$ in the base population.
The prediction error variance of the vector $\mathbf{\hat{g}}$ is usually 
globally obtained using cross validation. 
Alternatively they could be obtained individually by the posterior distribution 
of $(\mathbf{M} - 2\mathbf{p})\mathbf{\hat{u}}$ when a Bayesian model is used.

## Single Step Transciptomic prediction of Breeding Values
The breeding values in a mRNA prediction model are in general given by:
$$
\begin{equation}
	\label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}
$$

where $\boldsymbol{\hat{\alpha}}$ is the solution to a ridge regression model 
equaivalent to the GBLUP model but using mRNA covariates instead of 
marker genotypes.

Consider now the situation in which only a subset of the population has 
transcriptomic information but all have SNP genotypes.
We can take a similar route as in @fernando_class_2014 and impute missing
mRNA data.
The difference here is that &mdash; instead of predicting SNP covariates using 
pedigree relationship &mdash; we impute MRNA covariates using SNP or genomic 
relationships.
Let the subscript $1$ denote individuals for which SNP genotypes but no mRNA 
data are available.

>>>>

---
output: pdf_document
---
# Introduction
The integration of different sources of information for the prediction of 
breeding or production values is one of the major challenges in current genomic 
prediction research.
In the early years of genomic prediction, a 2-step procedure was the most 
intuitive choice for establishing a new selection scheme.
The first consists of conventionally estimating breeding values based on 
pedigree relationships and therefore the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (i.e. high 
number of offspring), enter the second step in which SNP genotypes are regressed 
against the conventional breeding values.
Those markers can then be used to predict genomic breeding values for solely 
genotyped and potentially very young individuals.
Considering the two sources of information &mdash; pedigree and high density SNP 
genotypes &mdash; the process can be described as follows:

@@

where $\mathbf{X}$ is a design matrix for fixed effects, $\mathbf{b}$ is a 
vector of fixed effects, $\mathbf{Z}$ is the design matrix for the random animal 
effects, which associate observations in $\mathbf{y}$ with elements in 
$\mathbf{A}$, $\mathbf{a}$ is the vector of random animal effects (=breeding 
values) and $\mathbf{e}$ is a vector of iid residual errors with expectation 
$0$ and variance $\sigma^2_e$.
The distribution of the random vector $\mathbf{a}$ is 
$MVN(0, \mathbf{A}\sigma^2_a)$.

### Second Step
In the second step, a subset of the individuals for which conventional breeding 
values have been estimated, is selected and enters the reference population for 
training a genomic prediction model.
That selection requires that the individual is genotyped and has 
$\text{PEV}(a_{\text{candidate}}) \leq \text{threshold}$.
As the breeding values from step one enter the genomic prediction model as 
phenotypes, one has to be careful about the residual error distribution of
such a model.
For instance, if the range of the prediction error variances is very broad in 
the reference population, the residual errors will not be identically 
distributed anymore, e.g. they should be weighted differently or estimated 
individually.
One popular approach to account for that problem is *Deregressing* the breeding 
values and weighting the residuals according to the prediction error variances 
[@Garrick2009].
Genomic breeding values for any animal in the population can then be obtained 
by:

$$
\begin{equation}
	\label{eq:gebv}
	\mathbf{\tilde{g}} =  (\mathbf{M} - 2\mathbf{p})\mathbf{\hat{u}},
\end{equation}
$$

where $\mathbf{p}$ is a row-vector of allele frequencies in the base population,
$\mathbf{M}$ is a matrix of marker covariates and $\mathbf{u}$ is a vector of
marker effects.
The prediction error variance of the vector $\mathbf{\tilde{g}}$ is usually 
globally obtained using cross validation. 
Alternatively they could be obtained individually by the posterior distribution 
of $(\mathbf{M} - 2\mathbf{p})\mathbf{\hat{u}}$ when a Bayesian model is used.

## Single Step Transciptomic prediction of Breeding Values
The breeding values in a mRNA prediction model are in general given by:
$$
\begin{equation}
	\label{eq:mrnaebv}
	\mathbf{\tilde{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}
$$

where $\boldsymbol{\hat{\alpha}}$ is the solution to a ridge regression model 
equaivalent to the GBLUP model but using mRNA covariates ($\mathbf{W}$) instead of 
marker genotypes.

Consider now the situation in which only a subset of the population has 
transcriptomic information but all have SNP genotypes.
We can take a similar route as in @fernando_class_2014 and impute missing
mRNA data.
The difference here is that &mdash; instead of predicting SNP covariates using 
pedigree relationship &mdash; we impute mRNA covariates using SNP or genomic 
relationships.
Let the subscript $1$ denote individuals for which SNP genotypes but no mRNA 
data are available.
Individuals with subscript $2$ have both, SNP and mRNA information.
The covariates in $\mathbf{W}$ are centered.
The vector $\mathbf{g}_1$ can be written as the sum of the conditional 
expectation given $\mathbf{g}_2$ and a residual error term:
$$
\begin{align}
	\label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{G_{12}}\mathbf{G_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{G_{12}}\mathbf{G_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon}
\end{align}
$$

The covariance matrix of $\boldsymbol{\epsilon}$ is $\mathbf{G}_{11} - \mathbf{G}_{12}\mathbf{G}_{22}^{-1}\mathbf{G}_{21} = (\mathbf{G}^{11})^{-1}$
[@legarra_relationship_2009].

The structure of the residual imputation/prediction error is known and can 
therefore be modelled.
The missing mRNA covariates can be predicted using the expectation of a 
multivariate normal random vector given correlated observations, which are 
related to the Best Linear Predictor.

### Extending the single step model

To this point, we have only taken the general idea behind single step genomic prediction and applied
it to the situation in which the breeding values are defined as the sum of mRNA effects plus an *imputation error*
term from using genomic relationships to impute missing mRNA data.
It might also be interesting to include individuals which have neither mRNA nor SNP data available,
but phenotypes.
The general mRNA model for that situation could be written as:

$$
\begin{equation}
\mathbf{y} = \mathbf{Xb} + \mathbf{W} \boldsymbol{\alpha} + \mathbf{U}_{Ped} \boldsymbol{\epsilon}_{Ped} 
+ \mathbf{U}_{SNP} \boldsymbol{\epsilon}_{SNP}+ \mathbf{e},
\end{equation}
$$

with

$$
\begin{equation}
\mathbf{X} = 
\begin{bmatrix}
  X_0 \\
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_0\hat{\mathbf{W}_0} \\
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U}_{Ped} = 
\begin{bmatrix}
  Z_0 \\
  0 \\
  0 
 \end{bmatrix},
 \mathbf{U}_{SNP} = 
\begin{bmatrix}
  0 \\
  Z_1 \\
  0 
 \end{bmatrix},
 \end{equation}
$$

The subscript 0 denotes animals with Pedigree information and neither
SNP nor mRNA data, 1 individuals with SNP but no mRNA data and 2 individuals with mRNA data.

Following the notation in @fernando_class_2014 the phenotypes can be untangled like this:

$$
\begin{align}
\begin{bmatrix}
  y_0 \\
  y_1 \\
  y_2 
 \end{bmatrix}
& =
 \begin{bmatrix}
  X_0 \\
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_0 & 0 & 0 \\
  0 & Z_1 & 0 \\
  0 & 0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  g_0 \\
  g_1 \\
  g_2 
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_0 \\
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_0 & 0 & 0 \\
  0 & Z_1 & 0 \\
  0 & 0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{A}_{02}\mathbf{A}_{22}^{-1}\mathbf{W}_2\boldsymbol{\alpha} + \boldsymbol{\epsilon}_{Ped} \\
  \mathbf{G}_{12}\mathbf{G}_{22}^{-1}\mathbf{W}_2\boldsymbol{\alpha} + \boldsymbol{\epsilon}_{SNP}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_0 \\
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_0 & 0 & 0 \\
  0 & Z_1 & 0 \\
  0 & 0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \hat{\mathbf{W}_0}\boldsymbol{\alpha} + \boldsymbol{\epsilon}_{Ped} \\
  \hat{\mathbf{W}_1}\boldsymbol{\alpha} + \boldsymbol{\epsilon}_{SNP} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}
$$

The breeding values from that model are:

$$
\begin{equation}
\tilde{\mathbf{g}} = 
 \begin{bmatrix}
  \hat{\mathbf{W}_0} \\
  \hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix}
 \hat{\boldsymbol{\alpha}}
 + 
 \begin{bmatrix}
  \mathbf{Z}_0 \\
  0 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}}_{Ped}
 + 
 \begin{bmatrix}
  0 \\
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}}_{SNP} 
 \end{equation}
 $$

When dealing with hybrid performance as phenotypes, we are interested in estimating
general combining abilitiy (GCA) for every line of a specific subpopulation.
In our case, the subpopulations comprise *dent* and *flint*.
Hence, our final model looks like this:

$$
\begin{align}
\mathbf{y} &= \mathbf{Xb} \\ 
  &+ \mathbf{Z}_{dent}\mathbf{W}_{dent} \boldsymbol{\alpha}_{dent} + \mathbf{Z}_{dent}\mathbf{U}_{dent_{Ped}} \boldsymbol{\epsilon}_{dent_{Ped}}
+ \mathbf{Z}_{dent}\mathbf{U}_{dent_{SNP}} \boldsymbol{\epsilon}_{dent_{SNP}} \\
  &+ \mathbf{Z}_{flint}\mathbf{W}_{flint} \boldsymbol{\alpha}_{flint} + \mathbf{Z}_{flint}\mathbf{U}_{flint_{Ped}} \boldsymbol{\epsilon}_{flint_{Ped}}
+ \mathbf{Z}_{flint}\mathbf{U}_{flint_{SNP}} \boldsymbol{\epsilon}_{flint_{SNP}} \\
  &+ \mathbf{e},
\end{align}
$$

where $\mathbf{Z}_{dent}$ and $\mathbf{Z}_{flint}$ are design matrices that map hybrids to their respective
parents in the dent and flint lines.
The obtained GCA effects for the inbred lines represent half their breeding values.
The predicted hybrid performance is:

$$
\begin{equation}
\tilde{\mathbf{h}} = \mathbf{Z}_{dent}\tilde{\mathbf{g}}_{dent} + \mathbf{Z}_{flint}\tilde{\mathbf{g}}_{flint} 
\end{equation}
$$

### 2016-09-21 reports/loocv_reduced.Rmd
ss-pred report, single doc with toc

## Predictive ability for non-imputed data
Peak at predictive abilities based on non-imputed predictor information and rank
the results by their predictive ability.
```{r Non-Imputed-Pred-Ability, cache = TRUE}
loocv_tbl_cap <- paste("Predictive abilities for the 'reduced' set of hybrids",
                       "using two different predictors and seven agronomic",
                       "traits. Results are ranked by predictive ability.")
loocv_clean %>%
  filter(Dent_NA_Fraction == 0, Flint_NA_Fraction == 0) %>%
  mutate(Predictor = as.factor(Predictor)) %>%
  droplevels() %>%
  group_by(Predictor, Trait) %>%
  summarize(Pred_Ability = cor(y, yhat)) %>%
  arrange(desc(Pred_Ability)) %>%
  rename("Predictive Ability" = Pred_Ability) %>%
  knitr::kable(caption = loocv_tbl_cap, booktabs = TRUE)
```

>>>>

## Predictive ability for non-imputed data
Peak at predictive abilities based on non-imputed predictor information and rank
the results by their predictive ability.
```{r Non-Imputed-Pred-Ability, cache = TRUE}
loocv_tbl_cap <- paste("Predictive abilities for the 'reduced' set of hybrids",
                       "using two different predictors and seven agronomic",
                       "traits. Results are ranked by predictive ability.")
loocv_clean %>%
  filter(Dent_NA_Fraction == 0, Flint_NA_Fraction == 0) %>%
  mutate(Predictor = as.factor(Predictor)) %>%
  droplevels() %>%
  group_by(Predictor, Trait) %>%
  summarize(Pred_Ability = round(cor(y, yhat), digits = 2)) %>%
  arrange(desc(Pred_Ability)) %>%
  rename("Predictive Ability" = Pred_Ability) %>%
  knitr::kable(caption = loocv_tbl_cap, booktabs = TRUE)
```

### 2016-09-21 reports/na_fraction_pheno_predability.Rmd
ss-pred report, single doc with toc

full_loocv_clean <- full_loocv %>%
  mutate(Trait = as.factor(Trait),
         Trait = fct_recode(Trait, 
                            "FAT" = "FETT",
                            "PRO" = "RPR",
                            "DMY" = "GTM",
                            "DMC" = "GTS",
                            "SUG" = "XZ"),
         Trait = fct_relevel(Trait, c("DMY", "DMC", "ADF", "FAT", "PRO", "STA",
                                      "SUG")),
         Residuals = y - yhat,
         Predictor = as.factor(as.character(Imputation)),
         Predictor = fct_recode(Predictor,
                                "mRNA" = "TRUE",
                                "SNP" = "FALSE")
  )
full_loocv_tbl_cap <- paste("Predictive abilities for the 'full' set of hybrids",
                            "using two different predictors and seven agronomic",
                            "traits. Results are ranked by predictive ability.")
full_loocv_clean %>%
  droplevels() %>%
  group_by(Predictor, Trait) %>%
  summarize(Pred_Ability = cor(y, yhat)) %>%
  arrange(desc(Pred_Ability)) %>%
  rename("Predictive Ability" = Pred_Ability) %>%
  knitr::kable(caption = full_loocv_tbl_cap, booktabs = TRUE)
```

>>>>

full_loocv_clean <- full_loocv %>%
  mutate(Trait = as.factor(Trait),
         Trait = fct_recode(Trait, 
                            "FAT" = "FETT",
                            "PRO" = "RPR",
                            "DMY" = "GTM",
                            "DMC" = "GTS",
                            "SUG" = "XZ"),
         Trait = fct_relevel(Trait, c("DMY", "DMC", "ADF", "FAT", "PRO", "STA",
                                      "SUG")),
         Residuals = y - yhat,
         Predictor = as.factor(as.character(Imputation)),
         Predictor = fct_recode(Predictor,
                                "mRNA" = "TRUE",
                                "SNP" = "FALSE")
  )
full_loocv_tbl_cap <- paste("Predictive abilities for the 'full' set of hybrids",
                            "using two different predictors and seven agronomic",
                            "traits. Results are ranked by predictive ability.")
full_loocv_clean %>%
  droplevels() %>%
  group_by(Predictor, Trait) %>%
  summarize(Pred_Ability = round(cor(y, yhat), digits = 2)) %>%
  arrange(desc(Pred_Ability)) %>%
  rename("Predictive Ability" = Pred_Ability) %>%
  knitr::kable(caption = full_loocv_tbl_cap, booktabs = TRUE)
```
